\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-04-27 forwards-and-backwards.tex
edits after meeting

{\bf Reproduction parameters:}
We assume that 
\begin{enumerate}
            \addtocounter{enumi}{2}
\item The function $F(x,m)$ satisfies 
    \begin{enumerate}
    \item $F(x,m)$ is locally Lipschitz in $m$;
    \item $F(x,m)$ is uniformly bounded above (but not necessarily below);
    \item for each fixed $m$, $\sup_{x\in\IR^d}\sup_{k\leq m}|F(x,k)|<\infty$;
    \end{enumerate}
\item The functions $r(x,m)$, $\gamma(x,m)$ have bounded first and second 
    derivatives in both arguments;
    \label{r_gamma_derivs_condition}
\item $\gamma(x,m)$ is uniformly bounded;
    \label{gamma_bounded_condition}
\item For each $f \in C_b^2(\IR^d)$,
    there is a $C_f$ such that
    $$
        |\gamma(x, \smooth{\gamma} \eta(x)) \theta \int 
        (r(y, \smooth{r} \eta(y)) f(y) - r(x, \smooth{r} \eta(x)) f(x))
        q_\theta(x, dy)| \le C_f (1 + |f(x)|)
    $$
    for all $x \in \IR^d$ and $\eta \in \measures$.
    Furthermore, $C_f$ only depends on the norm of the first two derivatives of $f$,
    i.e., 
    \[
        C_f = C(\sup_x \sup_{\|z\|=1} \max( \sum_i z_i \partial_{x_i} f(x), \sum_{ij} z_i z_j \partial_{x_i x_j} f(x))) .
    \] \label{gamma_B_condition}
\end{enumerate}
To keep expressions manageable, we shall also assume that 
\[
\mu_\theta(x) = r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N),
\]
that is, this expression is non-negative so that there is no need to take the maximum
	with zero in~(\ref{eqn:mu_defn}). (This is anyway true for 
	sufficiently large $\theta$.)
\end{assumptions}

Since we take bounded $f$,
for most situations the bound $C_f(1+f(x))$ in condition~\ref{gamma_B_condition} above
can be safely replaced simply by $C_f$;
however, this will be useful
in certain situations where we consider a sequence of $f$ with increasing upper bounds.
We now give two concrete situations in which condition~\ref{gamma_B_condition} is satisfied.

\begin{lemma}
    \label{lem:conditions on r}
    Assume that Conditions~\ref{def:model_setup} are satisfied,
    except for condition~\ref{gamma_B_condition}.
    If either
\begin{enumerate}
\item
\label{control through r} 
    $|\partial_{x_i}r(x,\smooth{r}\eta(x))|$ and 
    $|\partial_{x_ix_j}r(x,\smooth{r}\eta(x))|$
    are uniformly bounded for $x\in\IR^d$, $\eta\in\measures$;
\item 
\label{control through gamma}
    or, $m^2\gamma(x,m)$ is uniformly bounded and there exists $C<\infty$
    such that for $\theta$ sufficiently large, and all $x\in\IR^d$, $\eta\in\measures$,
    \[
    \theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)q_\theta(x,dy)\leq C\smooth{\gamma}\eta(x),
    \]
    and
    \[\theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)^2q_\theta(x,dy)
    \leq C(\smooth{\gamma}\eta(x))^2 ,
    \]
\end{enumerate} 
    then condition~\ref{gamma_B_condition} is also satisfied.
\end{lemma}

The purpose of the conditions that we have placed on the reproduction parameters is to 
ensure that the net per capita reproduction rate (before time scaling)
is order $1/\theta$. As remarked above, 
because of the non-local reproduction mechanism, it no longer suffices to assume that 
$r(x,\smooth{r}\eta(x))\gamma(x, \smooth{\gamma}\eta(x))-\mu_\theta(x)$ is of order $1/\theta$.
An instructive example is the case where $\gamma \equiv 1$ and $F \equiv 0$, so that $\mu_\theta = r$,
and $\eta = \delta_x$
(i.e., the population has all individuals at a single location),
in which case the mean rate of change of the total population size
is $\int (r(y, N \rho_r(y) ) - r(x, N \rho_r(x))) q_\theta(x, dy)$;
the first condition of Lemma~\ref{lem:conditions on r} would ensure this is of order $1/\theta$.

@@

\begin{proposition}
    \label{prop:nonlocal_to_local}
Let notation $\rho^\epsilon_F(x)=\rho_F\big(x/\epsilon)/\epsilon^d$.
Assume $\varphi_0\in L^2(\IR)$ is a positive, uniformly Lipschitz, and uniformly bounded function. 
Suppose that $\varphi^\epsilon\in L^2(\IR^d)$ is a weak solution to the equation
\begin{equation}
\label{nonlocalPDEv1} 
%\begin{cases}
\partial_t \varphi^\epsilon = \DG^* \varphi^\epsilon + 
\varphi^\epsilon F(\rho_F^\epsilon*\varphi^\epsilon),  
	%& 
\qquad x \in \mathbb{R}^d,\, t >0, %\\ 
%	\varphi^\epsilon(0,x) = \varphi_0(x) & x \in \mathbb{R}^d;
%\end{cases}
\end{equation}
with initial condition $\varphi_0(\cdot)$,
and that $\varphi$ is a weak solution to the equation
 \begin{equation}
\label{localPDE} 
%\begin{cases}
\partial_t \varphi = \DG^* \varphi + \varphi F(\varphi),  
%	& 
\qquad x \in \mathbb{R}^d,\, t >0, %\\ 
%	\varphi(0,x) = \varphi_0(x) & x \in \mathbb{R}^d.
%\end{cases}
\end{equation}
also with initial condition $\varphi_0(\cdot)$.
%Assume $\varphi_0\in L^2(\IR)$ is a positive, Lipschitz, and uniformly bounded function; 
Suppose further that $F$ is a Lipschitz function which is bounded above, and that
$\meanq(x)$ and $\covq(x)$, the drift and 
covariance matrix of $\DG$, satisfy the conditions of 
Assumptions~\ref{def:model_setup}.
Then, for all $T>0$ there exists a constant $K=K(T, \Vert \varphi_0 \Vert_\infty) < \infty$
and a function $\delta(\epsilon)$ (dependent on $\rho_F$) with $\delta(\epsilon)\to 0$ 
as $\epsilon\to 0$,
such that, for all $0 \leq t \leq T$, and $\epsilon$ small enough,
\[ 
\Vert \varphi_t(\cdot) - \varphi_t^\epsilon(\cdot) \Vert_\infty\leq K\delta(\epsilon). 
\]
In particular, as $\epsilon\to 0$, we have that $\varphi^\epsilon$ converges 
uniformly in compact intervals of time to $\varphi$.
\end{proposition}

@@

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^N_t)_{t \geq 0}$
    be as defined in Definition~\ref{defn:mgale_construction} with
$r(x,m)\equiv 1\equiv \gamma(r,m)$, $F(x,m)\equiv F(m)$, 
$\rho^\epsilon_F$ a symmetric Gaussian density with variance
parameter $\epsilon^2$,
and $\DG=\Delta$. 
Further suppose that $F(m)$ is a polynomial with $F(m)\ind_{m\geq 0}$ bounded above.
Assume that $\langle 1,\eta_0\rangle$ is bounded, and that for all 
$x\in\IR^d$ and $k\in \IN$,
	\[
		\limsup_{\epsilon\to 0} 
		\IE\big[\rho_F^\epsilon*\eta_0(x)^k\big]<\infty,
	\]
	and 
\[
    \limsup_{\epsilon \to 0} \int \IE\big[\rho_F^\epsilon*\eta_0(x)^k\big] dx<\infty.
\]
Finally assume that $N \to \infty$, $\theta \to \infty$
and $\epsilon \to 0$ in such a way that
\begin{equation}
\frac{1}{\theta\epsilon^2}+\frac{\theta}{N\epsilon^d}\to 0.
\end{equation}
Then the sequence of 
    ${\mathcal D}_{[0,\infty)}(\measures)$-valued
stochastic processes $\big(\rho_F^\epsilon*\eta^N_t(x)dx\big)_{t \ge 0}$
converges weakly
to a measure-valued process with a density $\varphi(t, x)$
that solves
\begin{align}
	\label{target equation in local convergence}
        \partial_t \varphi(t, x)=\Delta \varphi (t,x) +\varphi(t,x) F(\varphi(t,x)).
\end{align}
\end{theorem}

@@

\begin{corollary} \label{cor:wavefront}
    In addition to the assumptions of Corollary~\ref{cor:lineages_simple},
    suppose that the population process is described by a travelling wave with 
velocity $\wavespeed$,
    i.e., the population has density
    $\varphi(t, x) = w(x - t \wavespeed)$
    where $w$ solves
    \begin{align*}
        r \DG^* (\gamma w) + w F + \wavespeed \cdot \grad w = 
r\sigma^2\Delta (\gamma w) -\meanq \cdot\grad (\gamma w) +\wavespeed \cdot \grad w =0 .
    \end{align*}
    Then the semigroup $Q_s$
    of the motion of a lineages in frame that is moving at speed $\wavespeed$
    is time-homogeneous with generator
    \begin{align}
\label{Lgen}
        \Lgen f
        &=
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            2 \grad \log (\gamma w)
            \cdot \grad f
        \right)
        + (\wavespeed - r\gamma \meanq) \cdot \grad f .
    \end{align}
\end{corollary}

@@

Equation~(\ref{eqn:nonlocal_pde}) is a nonlocal version of a reaction-diffusion equation;
the diffusion is nonlinear if $\gamma$ depends on population density:
in other words, if the diffusivity of the population depends on the population density.
Passing to the classical limit, we recover equations like~(\ref{PME}).
Such equations are widely used in a number of contexts in biology in which
motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see~\cite{sherratt:2010} and references therein for further discussion. 
In particular, it has been suggested as a model for
the expansion of a certain type of bacteria % of the type {\em Paenbacillus dendritiformis}
on a thin layer of agar in a Petri dish 
\citep{cohen/golding/kozlovsky/benjacob/ron:1999}. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population. 
We focus instead on the monostable reaction of~(\ref{PME}).

Comparing %\eqref{PME} 
with \eqref{eqn:nonlocal_pde},
we see that to set up a limit in which the population density $\varphi$ follows the 
porous medium equation with logistic growth of~(\ref{PME}),
we need $r=1$,
$\gamma = \varphi$, and $F = 1 - \varphi$.
Consulting equation~\eqref{eqn:mu_defn},
this implies that $\mu_\theta = \max\big(0, (1 + 1/\theta) \varphi - 1/\theta\big)$.
In other words,
establishment is certain
and birth rates increase linearly with population density,
but to compensate, death rates increase slightly faster (also linearly).
Alert readers will notice that
the condition from Assumptions~\ref{def:model_setup} 
that $\gamma(x,m)$ be uniformly bounded 
is violated.
This can be corrected by use of a cut-off, and in fact the downwards drift
provided by the logistic control of the population size prevents $m$ from 
getting too big. 
In practice in our simulations shown in Figure~\ref{fig:pme_waves}
take discrete time steps of length $dt$ (with $dt$ suitably small),
and have each individual reproduce and die with probabilities, respectively,
\begin{align*}
    p_\text{birth}(m) = \left(1 - e^{- m dt}\right)
    \qquad 
    p_\text{death}(m) = \left(1 - e^{- (m (1+1/\theta) - 1/\theta) dt} \right) ,
\end{align*}
where $m$ is the local density at their location.
This makes $\gamma(x, m) = p_\text{birth}(m)/dt \approx m$
and
\begin{align*}
    F(x, m) = \theta(\gamma(x, m) - \mu(x, m)) / dt \approx 1 - m .
\end{align*}
Birth and death rates are equal at density $m = 1$,
corresponding to an unscaled density of $N$ individuals per unit area.

\begin{figure}
    \begin{center}
        % \includegraphics[width=0.7\textwidth]{figures/ex2a/pme_1234.steps}
        % \includegraphics[width=0.7\textwidth]{figures/ex2b/pme_1234.steps}
        \includegraphics[width=0.7\textwidth]{figures/fig2a}
        \includegraphics[width=0.7\textwidth]{figures/fig2b}
    \end{center}
    \caption{
        Simulated 1D populations under a porous medium equation with logistic growth~(\ref{PME}) in $d=1$,
        % comparing the simulated profile to the analytical solution;
        $\theta/N$ small on the left; large on the right.
        % Simulation is with $\gamma(m) = m$ and $\mu(m) = (1+1/\theta) m - 1/\theta$
        % (implemented by choosing a small $dt$ taking reproduction probability per time step
        % to be $1 - \exp(- dt m)$
        % and death probability to be $\mu(m) = 1 - \exp(-dt ((1+1/\theta) m - 1/\theta))$);
        Values of $\theta$ are 1 and 100, respectively, and both have $N$ set
        so that the density is roughly 100 individuals per unit of habitat
        (as displayed on the vertical axis).
        See text for details of the simulations.
        \label{fig:pme_waves}
    }
\end{figure}

@@

%%%%%%%%%%
\paragraph{Allen-Cahn equation:}
Now take the Allen-Cahn equation:
\begin{align} \label{eqn:allen_cahn}
    \partial_t\varphi = \partial_{xx}\varphi + \varphi(1-\varphi)(2\varphi-1+s),
\end{align}
for a given $s \in (0,2)$.
Once again we have taken $r=\gamma=1$, but now the reaction term
$F(x,m) = (1-m) (2m-1+s)$ is bistable.
This equation can be used to model the motion of so-called 
hybrid zones in population genetics; see, for example,
\citet{barton:1979}, \citet{gooding:2018}, and \citet{etheridge/gooding/letter:2022}.
This equation has an explicit travelling wave solution with speed $s$
and shape
\[ w^A(x) = (1+e^{x})^{-1} , \]
i.e., $\phi_t(x)=w^A(x-st)$ solves~(\ref{eqn:allen_cahn}).
Substituting $w^A$ in place of $w^F$ in~(\ref{eqn:fkpp_generator}),
we find that the generator of an ancestral lineage is now, 
\begin{align*}
\Lgen_A f
    &=
    \partial_{xx}f
    + 
    2 \frac{\partial_xw^A}{w^A} \partial_xf
    +
    s \partial_x f\\
    \qquad &=
    \partial_{xx}f
    -
    2 \frac{e^x}{1+e^x} \partial_xf 
    + 
    s \partial_xf,
\end{align*}
so lineages in the tip are pushed leftwards into the bulk of the 
wave at a rate $s-2e^x/(1+e^{x})$.
The strength of this drift decreases exponentially as $x\to -\infty$.
The density of the speed measure for this diffusion is
$$
    m_A(x) \propto e^{sx}(1+e^x)^{-2},
$$
which is integrable, and so determines the unique stationary distribution.
% (agreeing with Lemma~\ref{cor:stationary_dist}).
Thus the position of the ancestral lineage relative to the wavefront will converge to a stationary 
distribution which is maximised away from the extreme tip of the wave.
This is consistent with~\cite{etheridge/penington:2022}, who consider an analogous
stochastic population model, although the stronger result there (that the genealogy
of a sample from behind the wavefront is approximately a Kingman coalecsent) requires
the stronger condition $s<1$.

@@

More mathematically, consider the case in which
$\DG = \sigma^2 \Delta$ and all 
parameters are spatially homogeneous,
so that $r(x, \eta) = r(\smooth{r}\eta(x))$, and similarly for $\gamma$ and $F$.
If $\varphi_0$ is such that $F(\varphi_0)=0$ and $F'(\varphi_0)<0$, then the
constant solution $\varphi\equiv\varphi_0$ is a nontrivial equilibrium
of~\eqref{general deterministic limit}. However, it may be not unique and/or unstable,
and a stable solution may have oscillations on a scale determined by the interaction distance.

@@

Our characterisation of the motion of lines of descent (from which we establish
that of ancestral lineages) when we pass
to the scaling limit in our model will be justified via this construction.
In this section 
we present a lookdown construction for the general population model 
of Definition~\ref{defn:mgale_construction}. It will be
in the spirit of~\cite{kurtz/rodrigues:2011}. 
The general set-up is as follows.
Each individual will be labelled with a ``level'', a number in $[0, N]$.
We will still encode the process embellished by these levels
as a point measure:
if the $i^\mathrm{th}$ individual's spatial location is $x_i$
and level is $u_i$, then we will write
\[
    \lp^N = \sum_i \delta_{x_i, u_i} ,
\]
which is a measure on $\IR^d \times [0, N]$.
Note that each individual contributes mass 1 to the measure,
not $1/N$ as above. If we assign mass $1/N$ to each individual and 
ignore the levels we will recover our population model. 
Moreover, at any time, the levels of 
individuals in a given spatial region 
will be exchangeable and conditionally uniform on $[0, N]$:
in particular, choosing the $k$ individuals with the lowest levels 
in that region
is equivalent to taking a uniform random sample of size $k$ 
from the population in the region.
However, this exchangeability is only as regards the \emph{past}:
an individual's level encodes information about their future reproductive output,
since individuals with lower levels tend to live longer, and have more offspring.
For more explanation of the set-up and how this is possible,
see \citet{kurtz/rodrigues:2011} and \citet{etheridge/kurtz:2019} 
(and note that our $N$ corresponds to the $\lambda$ of those papers).
The power of this approach is that we can pass to a limit under the
same scalings as described in Theorem~\ref{thm:nonlocal_convergence}, 
and 
the limiting ``spatial-level'' process will still be a point measure,
and so we explicitly retains the notion of individuals and lineages
in the infinite-population limit.

>>>>

{\bf Reproduction parameters:}
We assume that 
\begin{enumerate}
            \addtocounter{enumi}{2}
\item The function $F(x,m)$ satisfies 
    \begin{enumerate}
    \item $F(x,m)$ is locally Lipschitz in $m$;
    \item $F(x,m)$ is uniformly bounded above (but not necessarily below);
    \item for each fixed $m$, $\sup_{x\in\IR^d}\sup_{k\leq m}|F(x,k)|<\infty$;
    \end{enumerate}
\item The functions $r(x,m)$, $\gamma(x,m)$ have bounded first and second 
    derivatives in both arguments;
    \label{r_gamma_derivs_condition}
\item $\gamma(x,m)$ is uniformly bounded;
    \label{gamma_bounded_condition}
\item For each $f \in C_b^2(\IR^d)$,
    there is a $C_f$ such that
    $$
        |\gamma(x, \eta) \theta \int 
        (r(y,\eta) f(y) - r(x, \eta) f(x))
        q_\theta(x, dy)| \le C_f (1 + |f(x)|)
    $$
    for all $x \in \IR^d$ and $\eta \in \measures$.
    Furthermore, $C_f$ only depends on the norm of the first two derivatives of $f$,
    i.e., 
    \[
        C_f = C(\sup_x \sup_{\|z\|=1} \max( \sum_i z_i \partial_{x_i} f(x), \sum_{ij} z_i z_j \partial_{x_i x_j} f(x))) .
    \] \label{gamma_B_condition}
\end{enumerate}
To keep expressions manageable, we shall also assume that 
\[
\mu_\theta(x) = r(x,\eta) \gamma(x,\eta) - \frac{1}{\theta} F(x,\eta),
\]
that is, this expression is non-negative so that there is no need to take the maximum
	with zero in~(\ref{eqn:mu_defn}). (This is anyway true for 
	sufficiently large $\theta$.)
\end{assumptions}

Since we take bounded $f$,
for most situations the bound $C_f(1+|f(x)|)$ in condition~\ref{gamma_B_condition} above
can be safely replaced simply by $C_f$;
however, this will be useful
in certain situations where we consider a sequence of $f$ with increasing upper bounds.
We now give two concrete situations in which condition~\ref{gamma_B_condition} is satisfied.

\begin{lemma}
    \label{lem:conditions on r}
    Assume that Conditions~\ref{def:model_setup} are satisfied,
    except for condition~\ref{gamma_B_condition}.
    If either
\begin{enumerate}
\item
\label{control through r} 
    $|\partial_{x_i}r(x,\eta)|$ and 
    $|\partial_{x_ix_j}r(x,\eta)|$
    are uniformly bounded for $x\in\IR^d$, $\eta\in\measures$;
\item 
\label{control through gamma}
    or, $m^2\gamma(x,m)$ is uniformly bounded and there exists $C<\infty$
    such that for $\theta$ sufficiently large, and all $x\in\IR^d$, $\eta\in\measures$,
    \[
    \theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)q_\theta(x,dy)\leq C\smooth{\gamma}\eta(x),
    \]
    and
    \[\theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)^2q_\theta(x,dy)
    \leq C(\smooth{\gamma}\eta(x))^2 ,
    \]
\end{enumerate} 
    then condition~\ref{gamma_B_condition} is also satisfied.
\end{lemma}

The purpose of the conditions that we have placed on the reproduction parameters is to 
ensure that the net per capita reproduction rate (before time scaling)
is order $1/\theta$. As remarked above, 
because of the non-local reproduction mechanism, it no longer suffices to assume that 
$r(x,\eta)\gamma(x,\eta)-\mu_\theta(x)$ is of order $1/\theta$.
Perhaps the simples example in which we can see this
is the case where $\gamma \equiv 1$ and $F \equiv 0$, so that $\mu_\theta = r$,
and $\eta = \delta_x$
(i.e., the population has all individuals at a single location),
so that $\smooth{r} \eta(y) = \rho_r(y)$.
In this case, the mean rate of change of the total population size
is $\int (r(y, \rho_r(y) ) - r(x, \rho_r(x))) q_\theta(x, dy)$;
the first condition of Lemma~\ref{lem:conditions on r} would ensure this is of order $1/\theta$.

@@

\begin{proposition}
    \label{prop:nonlocal_to_local}
Let $\rho^\epsilon_F(x)=\rho_F\big(x/\epsilon)/\epsilon^d$.
Assume $\varphi_0\in L^2(\IR)$ is a positive, uniformly Lipschitz, and uniformly bounded function. 
Suppose that $\varphi^\epsilon\in L^2(\IR^d)$ is a weak solution to the equation
\begin{equation}
\label{nonlocalPDEv1} 
%\begin{cases}
\partial_t \varphi^\epsilon = \DG^* \varphi^\epsilon + 
\varphi^\epsilon F(\rho_F^\epsilon*\varphi^\epsilon),  
	%& 
\qquad x \in \mathbb{R}^d,\, t >0, %\\ 
%	\varphi^\epsilon(0,x) = \varphi_0(x) & x \in \mathbb{R}^d;
%\end{cases}
\end{equation}
with initial condition $\varphi_0(\cdot)$,
and that $\varphi$ is a weak solution to the equation
 \begin{equation}
\label{localPDE} 
%\begin{cases}
\partial_t \varphi = \DG^* \varphi + \varphi F(\varphi),  
%	& 
\qquad x \in \mathbb{R}^d,\, t >0, %\\ 
%	\varphi(0,x) = \varphi_0(x) & x \in \mathbb{R}^d.
%\end{cases}
\end{equation}
also with initial condition $\varphi_0(\cdot)$.
%Assume $\varphi_0\in L^2(\IR)$ is a positive, Lipschitz, and uniformly bounded function; 
Suppose further that $F$ is a Lipschitz function which is bounded above, and that
$\meanq(x)$ and $\covq(x)$, the drift and 
covariance matrix of $\DG$, satisfy the conditions of 
Assumptions~\ref{def:model_setup}.
Then, for all $T>0$ there exists a constant $K=K(T, \Vert \varphi_0 \Vert_\infty) < \infty$
and a function $\delta(\epsilon)$ (dependent on $\rho_F$) with $\delta(\epsilon)\to 0$ 
as $\epsilon\to 0$,
such that, for all $0 \leq t \leq T$, and $\epsilon$ small enough,
\[ 
\Vert \varphi_t(\cdot) - \varphi_t^\epsilon(\cdot) \Vert_\infty\leq K\delta(\epsilon). 
\]
In particular, as $\epsilon\to 0$, we have that $\varphi^\epsilon$ converges 
uniformly in compact intervals of time to $\varphi$.
\end{proposition}

@@

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^N_t)_{t \geq 0}$
    be as defined in Definition~\ref{defn:mgale_construction} with
$r(x,m)\equiv 1\equiv \gamma(r,m)$, $F(x,m)\equiv F(m)$, 
$\rho^\epsilon_F$ a symmetric Gaussian density with variance
parameter $\epsilon^2$,
and $\DG=\Delta/2$. 
Further suppose that $F(m)$ is a polynomial with $F(m)\ind_{m\geq 0}$ bounded above.
Assume that $\langle 1,\eta_0^N\rangle$ is uniformly bounded, and that for all 
$x\in\IR^d$ and $k\in \IN$,
	\[
		\limsup_{\epsilon\to 0} 
		\IE\big[\rho_F^\epsilon*\eta_0(x)^k\big]<\infty,
	\]
	and 
\[
    \limsup_{\epsilon \to 0} \int \IE\big[\rho_F^\epsilon*\eta_0(x)^k\big] dx<\infty.
\]
Finally assume that $N \to \infty$, $\theta \to \infty$
and $\epsilon \to 0$ in such a way that
\begin{equation}
\frac{1}{\theta\epsilon^2}+\frac{\theta}{N\epsilon^d}\to 0.
\end{equation}
Then the sequence of 
    ${\mathcal D}_{[0,\infty)}(\measures)$-valued
stochastic processes $\big(\rho_F^\epsilon*\eta^N_t(x)dx\big)_{t \ge 0}$
converges weakly
to a measure-valued process with a density $\varphi(t, x)$
that solves
\begin{align}
	\label{target equation in local convergence}
        \partial_t \varphi(t,x) = \frac{1}{2} \Delta \varphi(t,x) +\varphi(t,x) F(\varphi(t,x)).
\end{align}
\end{theorem}

@@

\begin{corollary} \label{cor:wavefront}
    In addition to the assumptions of Corollary~\ref{cor:lineages_simple},
    suppose that the population process is described by a travelling wave with 
velocity $\wavespeed$,
    i.e., the population has density
    $\varphi(t, x) = w(x - t \wavespeed)$
    where $w$ solves
    \begin{align*}
        r \DG^* (\gamma w) + w F + \wavespeed \cdot \grad w = 
r\sigma^2\Delta (\gamma w) -\meanq \cdot\grad (\gamma w) +\wavespeed \cdot \grad w =0 .
    \end{align*}
    Then the semigroup $Q_s$
    of the motion of a lineage in the frame that is moving at speed $\wavespeed$
    is time-homogeneous with generator
    \begin{align}
\label{Lgen}
        \Lgen f
        &=
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            2 \grad \log (\gamma w)
            \cdot \grad f
        \right)
        + (\wavespeed - r\gamma \meanq) \cdot \grad f .
    \end{align}
\end{corollary}

@@

Equation~(\ref{eqn:nonlocal_pde}) is a nonlocal version of a reaction-diffusion equation;
the diffusion is nonlinear if $\gamma$ depends on population density:
in other words, if the diffusivity of the population depends on the population density.
Passing to the classical limit, we recover equations like~(\ref{PME}).
Such equations are widely used in a number of contexts in biology in which
motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see~\cite{sherratt:2010} and references therein for further discussion. 
In particular, it has been suggested as a model for
the expansion of a certain type of bacteria % of the type {\em Paenbacillus dendritiformis}
on a thin layer of agar in a Petri dish 
\citep{cohen/golding/kozlovsky/benjacob/ron:1999}. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population. 
We focus on the monostable reaction of~(\ref{PME}).

Comparing %\eqref{PME} 
with \eqref{eqn:nonlocal_pde},
we see that to set up a limit in which the population density $\varphi$ follows the 
porous medium equation with logistic growth of~(\ref{PME}),
we need $r=1$,
$\gamma = \varphi$, and $F = 1 - \varphi$.
Consulting equation~\eqref{eqn:mu_defn},
this implies that $\mu_\theta = \max\big(0, (1 + 1/\theta) \varphi - 1/\theta\big)$.
In other words,
establishment is certain
and birth rates increase linearly with population density,
but to compensate, death rates increase slightly faster (also linearly).
Alert readers will notice that
the condition from Assumptions~\ref{def:model_setup} 
that $\gamma(x,m)$ be uniformly bounded 
is violated.
This can be corrected by use of a cut-off, and in fact the downwards drift
provided by the logistic control of the population size prevents $m$ from 
getting too big. 
In practice the simulations shown in Figure~\ref{fig:pme_waves}
take discrete time steps of length $dt$ (with $dt$ suitably small),
and have each individual reproduce and die with probabilities, respectively,
\begin{align*}
    p_\text{birth}(m) = \left(1 - e^{- m dt}\right)
    \qquad 
    p_\text{death}(m) = \left(1 - e^{- (m (1+1/\theta) - 1/\theta) dt} \right) ,
\end{align*}
where $m$ is the local density at their location.
This makes $\gamma(x, m) = p_\text{birth}(m)/dt \approx m$
and
\begin{align*}
    F(x, m) = \theta(\gamma(x, m) - \mu(x, m)) / dt \approx 1 - m .
\end{align*}
Birth and death rates are equal at density $m = 1$,
corresponding to an unscaled density of $N$ individuals per unit area.

\begin{figure}
    \begin{center}
        % \includegraphics[width=0.7\textwidth]{figures/ex2a/pme_1234.steps}
        % \includegraphics[width=0.7\textwidth]{figures/ex2b/pme_1234.steps}
        \includegraphics[width=0.7\textwidth]{figures/fig2a}
        \includegraphics[width=0.7\textwidth]{figures/fig2b}
    \end{center}
    \caption{
        Simulated populations under a porous medium equation with logistic growth~(\ref{PME}) in $d=1$,
        % comparing the simulated profile to the analytical solution;
        $\theta/N$ small on the left; large on the right.
        % Simulation is with $\gamma(m) = m$ and $\mu(m) = (1+1/\theta) m - 1/\theta$
        % (implemented by choosing a small $dt$ taking reproduction probability per time step
        % to be $1 - \exp(- dt m)$
        % and death probability to be $\mu(m) = 1 - \exp(-dt ((1+1/\theta) m - 1/\theta))$);
        Values of $\theta$ are 1 and 100, respectively, and both have $N$ set
        so that the density is roughly 100 individuals per unit of habitat
        (as displayed on the vertical axis).
        See text for details of the simulations.
        \label{fig:pme_waves}
    }
\end{figure}

@@

%%%%%%%%%%
\paragraph{Allen-Cahn equation:}
Now take the Allen-Cahn equation:
\begin{align} \label{eqn:allen_cahn}
    \partial_t\varphi = \partial_{xx}\varphi + \varphi(1-\varphi)(2\varphi-1+s),
\end{align}
for a given $s \in (0,2)$.
Once again we have taken $r=\gamma=1$, but now the reaction term
$F(x,m) = (1-m) (2m-1+s)$ is bistable.
This equation can be used to model the motion of so-called 
hybrid zones in population genetics; see, for example,
\citet{barton:1979}, \citet{gooding:2018}, and \citet{etheridge/gooding/letter:2022}.
This equation has an explicit travelling wave solution with speed $s$
and shape
\[ w^A(x) = (1+e^{x})^{-1} , \]
i.e., $\phi_t(x)=w^A(x-st)$ solves~(\ref{eqn:allen_cahn}).
Substituting $w^A$ in place of $w^F$ in~(\ref{eqn:fkpp_generator}),
we find that the generator of an ancestral lineage relative to the wavefront is now, 
\begin{align*}
\Lgen_A f
    &=
    \partial_{xx}f
    + 
    2 \frac{\partial_xw^A}{w^A} \partial_xf
    +
    s \partial_x f\\
    \qquad &=
    \partial_{xx}f
    -
    2 \frac{e^x}{1+e^x} \partial_xf 
    + 
    s \partial_xf,
\end{align*}
so lineages in the tip are pushed leftwards into the bulk of the 
wave at a rate $s-2e^x/(1+e^{x})$.
The density of the speed measure for this diffusion is
$$
    m_A(x) \propto e^{sx}(1+e^x)^{-2},
$$
which is integrable, and so determines the unique stationary distribution.
% (agreeing with Lemma~\ref{cor:stationary_dist}).
Thus the position of the ancestral lineage relative to the wavefront will converge to a stationary 
distribution which is maximised away from the extreme tip of the wave.
This is consistent with~\cite{etheridge/penington:2022}, who consider an analogous
stochastic population model, although the stronger result there (that the genealogy
of a sample from behind the wavefront is approximately a Kingman coalecsent) requires
the stronger condition $s<1$.

@@

More mathematically, consider the case in which
$\DG = \sigma^2 \Delta$ and all 
parameters are spatially homogeneous,
so that $r(x, \eta) = r(\smooth{r}\eta(x))$, and similarly for $\gamma$ and $F$.
If $\varphi_0$ is such that $F(\varphi_0)=0$ and $F'(\varphi_0)<0$, then the
constant solution $\varphi\equiv\varphi_0$ is a nontrivial equilibrium
of~\eqref{general deterministic limit}.
However, this constant solution may not be unique, it may be unstable,
and a stable solution may have oscillations on a scale determined by the interaction distance.

@@

Our characterisation of the motion of lines of descent (from which we establish
that of ancestral lineages) when we pass
to the scaling limit in our model will be justified via a lookdown construction.
In this section 
we present such a construction for the general population model 
of Definition~\ref{defn:mgale_construction}. It will be
in the spirit of~\cite{kurtz/rodrigues:2011}. 
The general set-up is as follows.
Each individual will be labelled with a ``level'', a number in $[0, N]$.
We will still encode the process embellished by these levels
as a point measure:
if the $i^\mathrm{th}$ individual's spatial location is $x_i$
and level is $u_i$, then we will write
\[
    \lp^N = \sum_i \delta_{x_i, u_i} ,
\]
which is a measure on $\IR^d \times [0, N]$.
Note that each individual contributes mass 1 to the measure,
not $1/N$ as above. If we assign mass $1/N$ to each individual and 
ignore the levels we will recover our population model. 
Moreover, at any time, the levels of 
individuals in a given spatial region 
will be exchangeable and conditionally uniform on $[0, N]$:
in particular, choosing the $k$ individuals with the lowest levels 
in that region
is equivalent to taking a uniform random sample of size $k$ 
from the population in the region.
However, this exchangeability is only as regards the \emph{past}:
an individual's level encodes information about their future reproductive output,
since individuals with lower levels tend to live longer, and have more offspring.
For more explanation of the set-up and how this is possible,
see \citet{kurtz/rodrigues:2011} and \citet{etheridge/kurtz:2019} 
(and note that our $N$ corresponds to the $\lambda$ of those papers).
The power of this approach is that we can pass to a limit under the
same scalings as described in Theorem~\ref{thm:nonlocal_convergence}, 
and 
the limiting ``spatial-level'' process will still be a point measure,
and so we explicitly retain the notion of individuals and lineages
in the infinite-population limit.
